---
title: "CosMx Pancreas Analysis"
author: "Ash Fletcher"
date: "`r Sys.Date()`"
format:
  html:
    theme: cosmo
    code-copy: true
    code-link: true
    toc: true
    toc-depth: 3
    number-sections: true
---

# Install libraries

```{r}
#| label: install_packages
#| echo: true
#| eval: false
#| warning: false
#| message: false
# Install required packages if not already installed
install.packages("Seurat")
install.packages("tidyverse")
install.packages("ggplot2")
install.packages("future")
install.packages("BiocManager")
install.packages("openxlsx")
BiocManager::install("EnhancedVolcano")
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 8, fig.align = "center")
```

# Introduction

This document contains the analysis of CosMx spatial transcriptomics data from pancreatic tissue. 
The analysis includes data processing, visualization, and differential expression analysis.

# Load libraries

```{r libraries}
library(Seurat)
library(tidyverse)
library(openxlsx)
library(ggplot2)
library(future)
library(EnhancedVolcano)
# Sets working environment to hold larger data objects
options(future.globals.maxSize = 8000 * 1024^2)
```

# Plot colors

```{r helper_functions}
# Define color palette using glasbey and polychrome colors (uses Seurat package)
colors_20 <- DiscretePalette(20, palette = "glasbey", shuffle = FALSE)
colors_10 <- DiscretePalette(10, palette = "polychrome", shuffle = FALSE)
```

# Read in data- change `readRDS` to the path to where your data object is

```{r read data}
# This is a Seurat object
## Change path!
so_panc <- readRDS("~/Documents/GCB_Academy_2025/spatial_biology/cosmxsinglecell/so_panc_cosmxWTA_sct.rds") 
Idents(so_panc) <- "types"
```

# Basic Visualization

### Single FOV (field of view)

```{r}
#| label: single_fov
#| fig-width: 10
#| fig-height: 8
# Zoomed to single FOV: CellOverlay_F053.jpg
# Visualize cell types in zoom 1
ImageDimPlot(object = so_panc, fov = "zoom1", group.by = "types", axes = TRUE, cols = "glasbey")
```

### Multiple FOVs

```{r}
#| label: multiple_fovs
#| fig-width: 10
#| fig-height: 8
# Zoomed to 4 FOVs: CellOverlay_F051.jpg - CellOverlay_F054.jpg
# Visualize cell types in zoom 4
ImageDimPlot(object = so_panc, fov = "zoom4", group.by = "types", axes = TRUE, cols = "glasbey")
```

# Gene Expression Visualizations

### Assign marker genes to variables
```{r}
molsAll <- c("GCG", "EPCAM", "PRSS1", "C1QA",
          "ACTA2", "CD3E", "INS", "CD163", "CD86")
molsDuct <- c("KRT19", "SOX9", "CFTR", "MUC1")
molsBeta <- c("INS", "IAPP", "DLK1", "PDX1")
molsMacs <- c("CD68", "CD163", "CD86", "CSF1R", "MARCO")
molsPSC <- c("FAP", "ACTA2", "COL1A1", "PDGFRA")
```

### Ductal Markers

```{r}
#| label: ductal_markers
#| fig-width: 10
#| fig-height: 8
# Plot ductal markers (molecules) in single zoom segment
ImageDimPlot(so_panc, fov = "zoom1", cols = "polychrome", alpha = 0.25, molecules = molsDuct, 
             mols.size = 0.77, nmols = 20000, border.color = "black", coord.fixed = FALSE)
```

### Beta Cell Markers

```{r}
#| label: betacell_markers
#| fig-width: 10
#| fig-height: 8
# Plot Beta cell markers
ImageDimPlot(so_panc, fov = "zoom1", cols = "polychrome", alpha = 0.25, molecules = molsBeta, 
             mols.size = 0.77, nmols = 20000, border.color = "black", coord.fixed = FALSE)
```

### Macrophage Markers

```{r}
#| label: macrophage_markers
#| fig-width: 10
#| fig-height: 8
# Plot macrophage markers
ImageDimPlot(so_panc, fov = "zoom1", cols = "polychrome", alpha = 0.25, molecules = molsMacs, 
             mols.size = 0.77, nmols = 20000, border.color = "black", coord.fixed = FALSE)
```

### PSC (pancreatic stellate cell) Markers

```{r}
#| label: psc_markers
#| fig-width: 10
#| fig-height: 8
# Plot PSC markers- these are the fibroblasts of the pancreas
ImageDimPlot(so_panc, fov = "zoom1", cols = "polychrome", alpha = 0.25, molecules = molsPSC, 
             mols.size = 0.77, nmols = 20000, border.color = "black", coord.fixed = FALSE)
```

### All Cells Marker Genes

```{r}
#| label: molecule_visualization
#| fig-width: 10
#| fig-height: 8
# Plot a group of marker genes for all cell types
ImageDimPlot(so_panc, fov = "zoom1", cols = "polychrome", alpha = 0.25, molecules = molsAll, 
             mols.size = 0.35, nmols = 20000, border.color = "black", coord.fixed = FALSE)
```

## Cell Type Specific Genes in Specific Cells

```{r}
#| label: molecule_visualization_specific_cells
#| fig-width: 10
#| fig-height: 8
# Plot cell type specific genes in specific cells
ImageDimPlot(so_panc, fov = "zoom1", cells = WhichCells(so_panc, 
             # subset to the cells that you want to visualize
             idents = c("Acinar.1", "Acinar.2", "Ductal","Macrophage")), 
             cols = c("red", "green", "blue", "white"), 
             molecules = c("SLC4A4", "SYCN", "LYZ"),
             alpha = 0.3, mols.size = 0.77, mols.cols = c("cyan", "yellow", "purple"))
```

# Single-Cell Analysis: all cells in the dataset (all FOVs)

### UMAP Visualization - visualize clusters

```{r}
#| label: single_cell_analysis_clusters
#| fig-width: 10
#| fig-height: 8
# Visualize cluster UMAP
DimPlot(so_panc, group.by = "seurat_clusters") + scale_color_manual(values = colors_20)
```

### Visualize cell types

```{r}
#| label: single_cell_analysis_cells
#| fig-width: 10
#| fig-height: 8
# Visualize cell type UMAP
DimPlot(so_panc) + scale_color_manual(values = colors_20)
```

## Gene Expression Plots

### Violin plot

```{r}
#| label: gene_expression_violin
#| fig-width: 10
#| fig-height: 8
# Gene expression visualization
VlnPlot(so_panc, features = c("VIM", "PRSS1"), assay = "SCT", 
        layer = "counts", pt.size = 0.1) + NoLegend()
```

### Feature plot

```{r}
#| label: gene_expression_feature
#| fig-width: 10
#| fig-height: 8
FeaturePlot(so_panc, features = c("KRT17", "VIM", "CD163", "PRSS1"), 
            max.cutoff = "q95")
```

* What do you notice about the Acinar cell marker, PRSS1?


# Differential Expression Analysis

### Find Markers between Acinar.1 and Acinar.2: 
I already ran the `FindMarkers` algorithm in Seurat since it took me almost 10mins on my laptop.
Read in the differentially expressed genes table.
Change the path to where you downloaded the data to.

```{r}
#| label: differential_expression_find_markers
#| fig-width: 10
#| fig-height: 8
# Find markers between Acinar.1 and Acinar.2 ## This took me ~8mins on my laptop
# Since it takes so long, I provided the output in an Excel file
# 
# acini_markers <- FindMarkers(so_panc, ident.1 = "Acinar.1",
#                             ident.2 = "Acinar.2", verbose = TRUE, 
#                             assay = "SCT")

acini_markers <- openxlsx::read.xlsx("~/Documents/GCB_Academy_2025/spatial_biology/git_cosmx_analysis/Cosmx_WTA/acini_markers.xlsx", 
                    rowNames = TRUE) ### Change the path to where your excel file is!
```

### Volcano Plot of DEGs

```{r}
#| label: differential_expression_volcano
#| fig-width: 10
#| fig-height: 8
# Volcano plot parameters
fc <- 1
pval <- 0.05

degs <- acini_markers %>%
  filter(abs(avg_log2FC) > fc) %>% 
  filter(p_val_adj <= pval) 

# Create volcano plot
EnhancedVolcano(acini_markers,
                lab = rownames(acini_markers),
                selectLab = rownames(degs),
                subtitle = bquote(italic("Acinar.2 <---> Acinar.1")),
                x = 'avg_log2FC',
                y = 'p_val_adj',
                legendLabels = c("NS", paste0("Log2FC >", fc),
                                paste0("Adj. p-val <", pval), 
                                paste0("Adj. p-val <", pval, " and Log2FC >", fc)),
                subtitleLabSize = 15,
                captionLabSize = 15,
                pCutoff = pval,
                FCcutoff = fc,
                pointSize = 2.5,
                labSize = 4,
                axisLabSize = 14,
                legendLabSize = 14, 
                colAlpha = 0.35)
```

# Interactive Exploration

### See if/where your favorite genes show up in the pancreas

```{r}
#| label: gene_explore
#| fig-width: 10
#| fig-height: 8
my_genes <- c("SOX9") # Add up to 5 genes here

ImageDimPlot(so_panc, fov = "zoom4", # you can also change this to "zoom1"
             cols = "polychrome", alpha = 0.35, molecules = my_genes, 
             mols.size = 0.35, nmols = 20000, border.color = "black", coord.fixed = FALSE)
``` 
